---
title: "Drawing profile graphics for limnology and oceanography"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tdggraph_demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = 'center',
  fig.width = 4, 
  fig.height = 3,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
"Profile" data is common in both limnology and oceanography.  Profile data 
is data where some sort of water quality (or more rarely, biological) parameter
is measured from the water surface to depth. Often, profile data is collected
at regular intervals at the same location, producing a three dimensional 
(time by depth by value) data set.

Common examples of time by depth data include salinity, temperature and
dissolved oxygen profiles, but the idea applies whenever a parameter is
collected more or less simultaneously along a vertical line.

The goal of this package is to provide a simple interface to ggplot graphics,
with intelligent defaults to generate plots useful for this type of data.  
In particular, profile data is usually  drawn with a reversed vertical (depth) 
axis, so that the surface (zero depth) is at the top of the figure, with depth 
increasing towards the bottom of the figure. Similarly, it is common to show 
zero depth on the plot, even though data collection starts at some finite
depth below the water surface.

The `tdggraph` package is principally designed to produce graphics from 
season-long series of periodic profile observations. The three main graphics 
functions create `ggplot2` graphic objects with different strengths for 
emphasizing different aspects of the data.

# Dependencies 
We load several "tidyverse"  package individually here, to minimize
unnecessary dependencies in the package.  Users can usually load them in one 
step with `library(tidyverse)`.  
```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(ggplot2)

library(tdggraph)
```

# Load Example Data
We also create a factor version of the date to simplify examples.
```{r}
data(dep_sonde) 

dep_sonde <- dep_sonde %>%
  mutate(txt_date = format(sample_date, format = '%m/%d'),
         txt_date = factor(txt_date),
         txt_date = fct_reorder(txt_date, sample_date))
```

# `tdggraph` Principles
## Integration with `tidyverse` and `ggplot`
The functions in `tdggraph` are designed to integrate well with ggplot and dplyr 
work flows.  

The first argument to each plotting function is a source data frame.  They can
therefore be integrated into a `magrittr` pipe, like the `ggplot()`  All
graphic functions use data masking, so they will  look for data passed as 
function arguments within that data frame, before looking in the enclosing 
environment.  Variable names in the source data frame can therefore "mask" 
the same variable name from the enclosing environment.  This style of function
argument will be familiar to users of `ggplot()` and the tidyverse.

The output of each plotting function is of class `ggplot`, so users can add
additional graphic components to the plot as usual, adding layers, modifying 
themes, changing axis labels, altering color schemes, transforming axes, and 
more. 

The default reversal of the Y axis can be "undone" by adding 
`scale_y_continuous()` to the function output.  this will generate a warning,
but the Y axis will be returned to "normal" with higher values towards the top 
of the figure.

Users can pass additional aesthetic specifications as part of the 
call to any of these functions. Any additional arguments passed to these 
functions are passed directly to the underlying `ggplot2` geoms, allowing 
users to override aesthetic defaults, most of which are inherited from 
`ggplot()`. Additional arguments are passed to geoms as follows:

*  `ptlines()`  -> `geom_path()`  
*  `ptdots()`   -> `geom_dots()`  
*  `ptsmooth()` -> `geom_tile()`  

Passing a data dependent aesthetic specification using `aes()` requires some
special handling. Data masking does not work within the supplementary call to
`aes()`, so users must pass arguments to `aes()` that refer to objects defined 
in the enclosing environment.  Usually, this just requires an explicit reference
to the source data frame (e.g., `df$value`, not just `value`). See the example 
in the `ptdots()` "Filled Symbols" demo for an example.

## All Data From the Enclosing Environment
If all data is available in the enclosing environment, you can pass NULL as the 
data frame parameter.  This is sometimes helpful if you are using the functions 
inside another function, where you may not want to assemble a data frame.

## Order of Arguments
In the current version of the package, all plotting functions take arguments in
order of their role in the plot. The order follows the **graphical** logic of 
the plot, not the **scientific** logic. This is a potential source of errors and
frustration, so it's worth explaining.

The order of the arguments for the plot functions is always:  
*  data source;  
*  variable that defines x location;  
*  variable that defines y location;  
*  variable that defines color. 

From the scientific perspective, all these plots show depth, time, and a 
measured environmental variable. Those roles do not change, but the way they are
depicted on a plot does.

The formal arguments for `ptlines()` are as follows:
```{r eval = FALSE}
ptlines(.dt, .val, .y, .grp,.sort = TRUE, ...)
```
Note that in this type of plot the x coordinate reflects the value of the 
data you are showing, so it comes **first**.  The grouping variable defines
the color, so it comes **last**.

The first few formal arguments for `ptdots()` and `ptsmooth()` are similar to
each other:
```{r eval = FALSE}
ptdots(.dt, .x, .y, .val, ...)

ptsmooth(.dt, .x, .y, .val,
  .res_x = 0.5, .res_y = 2, 
  y_grow_grid = TRUE, y_with_zero = TRUE
)
```

The data source comes first, then the x coordinate (usually a time coordinate),
then the y coordinate (usually depth) then the value that defines the color 
(the measured environmental variable).



# `ptlines()` Demo
the `ptlines()` function generates a line graph with vertical profiles shown as 
lines.  Different profiles are distinguished by color.  This is an excellent
format to show subtle differences in profiles between a relatively modest number
of samples, as viewers  can easily compare profiles against each other.  It is
not as good if there are many samples, as a large number of lines begins to look 
like spaghetti.

This function can also be used to plot profile data from multiple sites, as the 
logic of the plot works with qualitative, not quantitative grouping variables.

Data is preferentially found inside the data frame passed as the first argument.
Arguments are in the order of x coordinate -- here a **value** , y coordinate --
usually the depth -- and a grouping variable -- often a date or a location.  The
grouping variable should evaluate to a factor or character vector, as it will
always be used to divide data into separate profiles.
```{r}
ptlines(dep_sonde, temp, depth, txt_date, .sort = FALSE)
```

## Integration with `tidyverse` and `ggplot`
The function is designed to fit into  a typical tidyverse work flow, for example, 
you can feed data into it via  a pipe, and add components to the resulting 
ggplot object:
```{r}
ptlines(dep_sonde, do, depth, txt_date, size = 1.5, lty = 2) +
  theme_minimal() +
  scale_color_brewer(type = 'qual', name = '') +
  ylab('Depth (m)') +
  xlab('Dissolved Oxygen (mg/l)')
```

Any supplementary aesthetics defined in `mapping = aes()` must refer to names 
from the enclosing environment.  See the example in the `ptdots()` "Filled 
Symbols" demo for an example.

## Data From the Enclosing Environment
If all data is available in the enclosing environment, you can pass NULL as the 
data frame parameter.  This is helpful if you are using the functions inside 
another function, where you may not want to assemble a data frame.
```{r}
td <- dep_sonde$txt_date
d = dep_sonde$depth
ox = dep_sonde$do

ptlines(NULL, ox, d, td, .sort = FALSE) +
  theme_minimal()
```

## Data Order Matters
Under the hood this function uses `geom_path()` and not `geom_line()`, so
the order of the data matters. `geom_path()` connects points in the order of
their occurrence in the source data, while geom_line() internally orders data
according to the x coordinate before plotting.

The DEP sonde data is already sorted by date and depth, whcih is why we were
able to ignore this detail up until now.  We demonstrate what happens by 
desorting the data we just plotted. By sorting each component of the data to the 
same random order, we ensure we are plotting the same data (x,y,z) triplets, 
only they are presented in a new random order.

You get spaghetti:
```{r}
rand_order <- order(rnorm(length(td)))
td2 <- td[rand_order]
d2 <- d[rand_order]
ox2 <- ox[rand_order]
ptlines(NULL, ox2, d2, td2, .sort = FALSE) +
  theme_minimal()
```

### The `.sort` Argument

it is often going to be a good practice to sort your data yourself before
passing it to this function, but if you're in a hurry, there's a shortcut that 
does the job for you.  You  control this behavior with the `.sort'
parameter, which is TRUE by default.  But be forwarned results may not quite
be what you expected or hoped for.
```{r}
ptlines(NULL, ox2, d2, td2, .sort = TRUE) +
  theme_minimal()
```

## Passing a Grouping Variable
Sometimes this gets tricky.  For example, suppose you want to symbolize
the data by month, not by date. You have multiple sampling dates within some
months.  By default, `geom_path()` will connect all the dots in the order of
the source data or the data sorted by depth.so neither `.sort = TRUE` or 
`.sort = FALSE` works quite as expected:

With `.sort = TRUE`, you get zigzags:
```{r}
ptlines(dep_sonde, temp, depth, month, .sort = TRUE)
```
With `.sort = FALSE`, things look better, because the source data was already 
sorted, but you get unneccesary connecting lines from the bottom of one profile 
to the tope of the next.
```{r}
ptlines(dep_sonde, temp, depth, month, .sort = FALSE)
```

The solution is to pass a grouping variable. Note: this behaves strangely with 
`.sort = TRUE`, so you should sort your data before passing it to `ptlines()`.
```{r fig.height = 3, fig,width = 4}
ptlines(dep_sonde, temp, depth, month,  .sort = FALSE,
        mapping = aes(group = dep_sonde$txt_date))
```

# `ptdots()` Demo
The dots provide an alternative way of depicting similar data.  Here, the
graphic emphasizes dates over the course of a season, at the expense
of making subtle differences in profiles more difficult to assess.  Default 
colors from `ggplot2` are often a poor choice here. 
```{r}
ptdots(dep_sonde, sample_date, depth, do, size = 5) +
  theme_minimal() +
  scale_color_gradient2(name = 'Oxygen, (mg/l)',
                        high = scales::muted("blue"), 
                        low = scales::muted("red"),
                        mid = 'gray95',
                        midpoint = 9)
```

## Data from the enclosing environment
As before, we can pass NULL as a data frame parameter if all data is available
in the enclosing environment.
```{r}
sd <- dep_sonde$sample_date
ptdots(NULL, sd, d, ox, size = 5) +
  theme_minimal()
```

## Filled Symbols
Any supplementary aesthetics defined in `mapping = aes()` must refer to names 
from the enclosing environment. As a result, using  filled point symbols. (R 
point shapes 21 through 25) takes several additional steps:

1. Add `mapping = aes(fill = ?)` to the function call, pointing 
   to data in the enclosing environment, usually by qualifying the name of 
   the data frame explicitly with `$`. 

2. Pass a `shape` argument selecting the filled symbol shape
  
3. Usually, pass a `color` argument to define the fixed dot outline color.
   (`color = 'gray50'` is often a good starting point).  If you don't specify
   a single color, to override the normal color aesthetic, the outline color 
   will vary,  which is probably not what you want. 

4. Usually, you will want to supplement the call to this function with a 
   call to `scale_fill_viridis_c()`, `scale_fill_distiller()`, or 
   `scale_fill_continuous()` or `scale_fill_gradient()` etc. to define a fill 
   color scale that works better than the `ggplot2` default.

```{r}
ptdots(dep_sonde, sample_date, depth, do, size = 3, 
       shape = 22, color = 'gray25',
       mapping = aes(fill = dep_sonde$do)) +
  theme_minimal() +
  scale_fill_gradient(name = 'Oxygen, (mg/l)',
                        high = "paleturquoise1", 
                        low = scales::muted("darkgreen")) +
  xlab('') +
  ylab('Depth (m)')
```

#`ptsmooth()` demo
This function produces a two dimensional interpolation plot. it can be thought of 
as a smoothed out version of `ptdots()`. It is especially useful for 
"eyeballing" seasonal patterns from periodic data. As before, default colors 
are seldom the most informative.
```{r}
data(dep_sonde)
plt <- ptsmooth(dep_sonde, sample_date, depth, temp, 
         .res_x = 1, .res_y = .25, 
         y_grow_grid = TRUE,
         y_with_zero = TRUE) +
scale_fill_gradient(name = 'Oxygen, (mg/l)',
                        high = scales::muted("blue"), 
                        low = scales::muted("red", l = 40))
plt
```

It is easy to overlay the locations of the original observations if you want to
show viewers where you have actual observations
```{r}
plt +
  geom_point(mapping = aes(sample_date, depth), data = dep_sonde,
             shape = 21, size = 3, fill = NA)
```

# `interpol()` Demo
This function is a thin wrapper around base R's `approx()` function, which 
implements simple linear interpolation.  The major difference is that the return 
value of `interpol()` is a data frame, not a list.  The function is used "under
the hood" to generate graphics interpolated in two directions without clipping.

The basic idea for this function and its use in generating smoothed graphics for
profile data  were inspired by Dewey Dunnington's 
fishandwhistle.net [blog](https://fishandwhistle.net/post/2019/depth-time-heatmaps/).

## By Depths / Numeric Values
```{r}
dat <- dep_sonde[dep_sonde$month == 'Jul',]
grid <- seq(0, 14, 0.75)
interp <- interpol(dat$depth, dat$temp, .grid = grid)

ggplot(dat, aes(temp, depth)) +
  geom_point(size = 3, col = 'red') +
  geom_point(mapping = aes(dep, ind), data = interp, shape = 3) +
  scale_y_reverse()
```

## By Dates / Times
```{r}
tmp <- dep_sonde %>% 
  select(site, sample_date, depth, month, temp) %>%
  group_by(sample_date) %>%
  filter(depth == min(depth)) %>%
  ungroup(sample_date)

grid <- seq(min(tmp$sample_date), max(tmp$sample_date), 7)

interp <- interpol(tmp$sample_date, tmp$temp, .grid = grid, 
                   grow_grid = FALSE,
                   na.rm = TRUE)

ggplot(tmp, aes(temp, sample_date)) +
  geom_point(size = 3, col = 'red') +
  geom_point(mapping = aes(dep, ind), data = interp, shape = 3) +
  ggtitle('Fore River Site 09, July 2018')
```

### Functional Programming
`interpol() includes an optional `.name` argument, which can be useful when
assembling larger data frames through workign with nested tibbles or functional
programming.  For example,  one might often want to interpolate within dates,
not between dates.

We calculate a grid to use for each date, then use a nested tibble to
interpolate within each date before using `reduce()` to generate a tidy
(long-form) tibble containing the results of the interpolation.
```{r}
lims <- range(dep_sonde$depth) 
max_y = ceiling(lims[2])
min_y = floor(lims[1])
ygrid = seq(min_y, max_y, by = 0.5)

profs <- dep_sonde %>%
    group_by(sample_date) %>%
    nest() %>%
    mutate(prof = map(data, function(dat) interpol(dat$depth, dat$do, 
                                                   .name = sample_date,
                                                   .grid = ygrid)))
profs <- reduce(profs$prof, bind_rows) %>%
  mutate(txt_date = factor(format(id, format = '%m/%d')),
         txt_date = fct_reorder(txt_date, id)) %>%
  rename(do = dep, depth = ind)
```

```{r}
ggplot(profs, mapping = aes(x = do, y = depth)) +
  geom_point(aes(color = txt_date))+
  geom_path(aes(color = txt_date)) +
  geom_point(data = dep_sonde, mapping = aes(x = do, y = depth))
```

## Passing an Empty Dataframe
Should return a data frame with NA in second variable.  This is useful for use
in functional programming, since it correctly propagates missing values.
```{r}
t <- unique(dep_sonde$sample_date) %>%
  sort
df <- tibble(x = t, y = NA_integer_)
interp <- interpol(df$x, df$y, .grid = grid, with_zero = FALSE, 
                   grow_grid = FALSE, na.rm = TRUE)
head(interp)
```

# `interpol_res()` Demo
`interpol_res()` is an only slightly more consequential extension of base R's
`approx()` function.  It's primary claim to fame is that it allows the user to
specify a resolution for evenly spaced interpolations. (`approx()` allows you to
specify the number of interpolated points, but not their spacing).   Like 
`interpol()`, it outputs a dataframe, not a list.

## By Depths / Numeric Values
```{r}
data(dep_sonde)
dat <- dep_sonde[dep_sonde$month == 'Jul',]
interp <- interpol_res(dat$depth, dat$temp, .res = 0.5)

ggplot(dat, aes(temp, depth)) +
  geom_point(size = 3, col = 'red') +
  geom_point(mapping = aes(dep, ind), data = interp, shape = 3) +
  scale_y_reverse()
```

## By Dates / Times
```{r}
tmp <- dep_sonde %>% 
  select(site, sample_date, depth, month, temp) %>%
  group_by(sample_date) %>%
  filter(depth == min(depth)) %>%
  ungroup(sample_date)
  
interp <- interpol_res(tmp$sample_date, tmp$temp, .res = 5, 
                   with_zero = FALSE, na.rm = TRUE)

ggplot(tmp, aes(temp, sample_date)) +
  geom_point(size = 3, col = 'red') +
  geom_point(mapping = aes(dep, ind), data = interp, shape = 3) +
  ggtitle('Fore River Site 09, July 2018')
```


# Gotchas
## Altering teh Vertical Scale on a "Reversed" Y Axis
Since the functions already define a "reversed" y axis, you need to be explicit
altering the y axis again. Sometimes this leads to unexpected results.

For example, when you alter the limits of the y axis, you need to provide 
the scale limits in reversed order (the way they are plotted). This is true 
both for `scale_y_reversed()` and for `coord_cartesian()`.  Note that these two 
commands do not quite generate identical plots, as `coord_cartesian()` will not
prune data.

```{r fig.height = 2, fig.width = 3}
plt <- ptdots(dep_sonde, sample_date, depth, do, size = 3)

plt + scale_y_reverse(limits = c(10, 0))

plt + coord_cartesian(ylim = c(10,0))
```

# Known Problems
## Functions Do Not Work with `facet_wrap()`


